FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-21 ess_marmo.tex
MATISSE description

\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay,
Orsay, France}
\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center,
Flagstaff, AZ, USA}
\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
CNRS, Sorbonne Universit\'es, UPMC Univ. Paris 06, Univ. Paris Diderot,
Sorbonne Paris Cit\'e, Meudon, France}
\affiliation{4}{Jacobs University, Bremen, Germany}
\affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
\affiliation{6}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
\affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.). Planetary missions produce an impressively growing 
amount of diverse data requiring an evolution from a mostly manual-based visual 
analysis to a more automated and quantitative analysis.
Interoperability, openness of data formats, and shared processing techniques are
becoming a necessity to efficiently extract scientific information from the data and 
to guarantee the reproducibility of the scientific results.
Unfortunately, the technologies and data formats used by researchers for
planetary surface studies and the field of astronomy 
diverged as these related, but almost completely isolated domains evolved.  
% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
% 
% > Unfortunately, the technologies 
% > and data formats used by researchers for planetary surface studies and the field of astronomy 
% > diverged as these related, but almost completely isolated domains evolved.  
% 
% Is this too strong a statement?
% 
% ^.
In this paper we will will describe how a small addition to the Flexible Image Transport System (FITS) 
standard, widely used in the astronomy investigations,  will allow FITS to be more easily used in planetary
surface investigations , we will  show  how FITS metadata can easily be transformed in the PDS4 metadata
archival model, and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from data acquisition to final visualization.
\end{abstract}

@@

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open and flexible, largely implemented in open and efficient
processing tools.
Those reasons make FITS format appropriate to automatically handle large amounts of raster data.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary surface domain could potentially
homogenize methods from acquisition, to visualization, while
giving chances to optimize data processing.

This paper is organized as follows:
\begin{enumerate}
\item{a brief description of FITS format;}
\item{a proposition for planetary surface metadata description in FITS;}
\item{dictionaries between FITS, PDS metadata description and EPN-TAP2.0
parameters;}
\item{some examples of planetary dataset in FITS;}
\item{current developments for making FITS more usable to handle data
of planetary surfaces.}
\end{enumerate}

@@

\subsection{FITS Format}
FITS is an open digital standard, defined for data acquisition and archiving
in astronomical observatories in the late 70's, and it is used for
spatial telescope data.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

FITS is already able to propose standard formatting for some data products
quite common by now in planetary surface investigations.
In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data,
vector, table data, etc.) in the same file, each with corresponding metadata,
as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
or hyperspectral cubes with geometry information (e.g. from CRISM
\citep{crism} or OMEGA \citep{omega} instruments).
FITS has been already chosen to distribute data from, e.g.,
Hayabusa AMICA and NIRS cameras\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
all Akatsuki cameras\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
(except the Lightning and Airglow Camera),
and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
In addition, many PDS3 datasets in the Small Bodies Node include FITS files.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by
the IAU Working Group on Cartographic Coordinates and Rotational Elements
(WGCCRE).
This group reports triennially \citep{wgccre} on the preferred rotation rate,
spin axis, prime meridian, and reference surface for planets and satellites
ensuring that cartographic endeavors are effectively comparable.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
project an extension to FITS metadata (GeoFITS) has been
proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
taking into account the WGCCRE recommendations \citep{MarmoLPSC}.

@@

FITS World Coordinate System
representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
way to describe spatial dependencies in FITS metadata.
WCS simplifies the spatial coordinate description with respect to historical
terrestrial references.
F
or instance, there is no need for oblique projection definitions
\citep{calagreisenI,Snyder1987} as projection parameters
(projection center or reference point)
describe the difference between, e.g., the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
This simplification is effective when a spherical local approximation
is applied on planetary imagery.
This is particularly the case with growing image resolutions and
increasing computing capabilities: spherical projections with a local
radius definition are easier to reproject from one system to another, than
from or to a global spheroidal datum.
As an example, the HiRISE team has chosen to distribute HiRISE projected
products using Equirectangular projection (CAR) on a local surface, approximated
by the sphere with radius equivalent to the local Mars spheroid radius.

In addition, WCS resolves the ambiguity between east-handed and west-handed
longitude systems, as the pixel conversion in world coordinates is set by an
oriented matrix.
If needed, FITS does offer the possibility to fully describe the projected
coordinate system using several alternative WCS definitions (meters, degrees,
East and West longitude, etc.)\citep{greisencalaII}.

Table \ref{table:proj} summarizes the FITS projection codes and their
corresponding PDS4 projection names (see table A.1. in \citet{calagreisenI}
for details).

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: the corresponding PDS4 terminology is
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
\end{tabular}
\end{table}

@@

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined (defaulted to \texttt{'ICRS'})}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed (defaulted to \texttt{'J2000'})}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in “4–3” form: the first four characters
specify the coordinate type, the fifth character is a '-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
are listed in Table \ref{table:ctype}.
For linear coordinates we propose a similar scheme: xPX as Projected X, xPY as Projected Y,
or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in degrees
and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
attributes.
In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
\item{CUNITn.
Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
and to meters (m) for linear ones.}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

\begin{table}
\label{table:ctype}
\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
Codes for planets are assumed in \citet{calagreisenI}. We propose to add codes for classes of targets,
more or less corresponding to PDS4 Target type attribute and to EPN-TAP target\_class parameter.}
\centering
\begin{tabular}{l l}
\hline
Body & CTYPE$^{a}$ \\ 
\hline
Moon & SE \\
Venus & VE \\
Mars & MA \\
Jupiter & JU \\
Saturn & SA \\
Uranus & UR \\
Neptune & NE \\
Satellites (other than the Moon) & ST \\
Asteroids & AS \\
Dwarf Planets & DW \\
Comets & CO \\
\hline
\multicolumn{2}{l}{$^{a}$From \citet{calagreisenI}}
\end{tabular}
\end{table}

@@

\subsubsection{Correspondence Between Reserved FITS Keywords and PDS4 Standard Reference}
\label{sssec:stdfitspds4}
The correspondence between mandatory and reserved FITS keywords (as defined in
the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
and PDS4 attributes, is described here.
\begin{enumerate}
\item{\texttt{BITPIX}.
It specifies the number of bits that represent a data value in
the associated data array.
PDS4 uses the attribute \texttt{data\_type}: corresponding 
values depend on the associated object class.}

@@

\item{\texttt{NAXISn}.
It contains a non-negative integer representing the number of
elements along axis \texttt{n} of a data array.
The \texttt{elements} attribute of the PDS4 \texttt{Axis\_Array} class provides
the count of the number of elements along an array.}

\item{\texttt{BUNIT}.
It contains a string, describing the physical units in which the quantities
in the array are expressed.
The units of all FITS header keyword values should conform with the recommendations
in the IAU Style Manual\footnote{https://www.iau.org/publications/proceedings\_rules/units/}.
For angular measurements given as floating point values, degrees are the recommended
units (with the units, if specified, given as 'deg').
PDS4 specifies in the class Element (of a vector, array, or any other class
containing Elements), a \texttt{unit\_id}} (unit abbreviation) and a \texttt{unit}
(unit long name).

\item{\texttt{BSCALE}.
It contains a floating point number representing the coefficient of the linear term
in the scaling equation $physical\_value = BZERO + BSCALE \times array\_value$,
the ratio of physical value to array value at zero offset.
For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
attribute.
Both FITS and PDS4 default this value to $1$.}

\item{\texttt{BZERO}.
It contains a floating point number representing the constant term in the
scaling equation $physical\_value = BZERO + BSCALE \times array\_value$
For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
attribute. Both FITS and PDS4 default this value to $0$.}

\item{\texttt{BLANK}.
This keyword must be used only with integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.
On floating point data the IEEE NaN must be used to represent undefined values.  
\texttt{PDS4 Special\_Constants?}}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UTC.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}

\item{\texttt{INSTRUME}.
It contains a character string identifying the instrument used to acquire the data.
PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally 
corresponds to this FITS keyword.
EPN-TAP provides with an \texttt{instrument\_name} parameter.}

\item{\texttt{TELESCOP}.
It contains a character string identifying the telescope used to acquire the data.
The PDS4 \texttt{Instrument\_Host} class provides the description of the physical object
mounting the instrument. Its \texttt{Name} attribute should correspond to the FITS keyword.
PDS4 also provides for the definition of a \texttt{Telescope} class, in case of
ground-based observations.
In the EPN-TAP scheme this is introduced by the \texttt{instrument\_host\_name}
parameter\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}.}

@@

\item{\texttt{ORIGIN}.
It contains string identifying the organization or institution responsible for
creating the FITS file.}

%
%\item{\texttt{AUTHOR}}

\item{\texttt{REFERENC}.
It contains a character string citing the bibliographic reference where the
data are first described.
It is recommended that either the Astrophysics Data System bibliographic
databases\footnote{http://adswww.harvard.edu/}
bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
be used.
PDS4 \texttt{External\_Reference} class and EPN-TAP \texttt{bib\_reference} have
exactly the same meaning and use than the FITS keyword.}

@@

\subsubsection{Visualization of Global and Projected Maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Given some new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
and GeoTIFF are completely interchangeable.
As an example we have downloaded some global maps from USGS
Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
We used ds9 to display the FITS image and were able to visualize coordinates in meters
as in QGIS, a popular open source GIS software, but also in planetary latitude and longitude
thanks to the alternate coordinate
description (see figure \ref{fig:ds9qgis}, left side).
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2}\includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9;
right: GeoTIFF raster superposed to the virtual
description of the FITS raster.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

Assuming projected products are distributed in FITS format, a quick method to add GIS support for
FITS images is to use a GDAL-supported detached header.
GDAL Virtual Header files simply describe the internal structure of a FITS image, but
also the map projection and body size in a standardized "well known text"
(WKT) projection string.
As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual Header
is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).

In order to load FITS files in GIS software, it is recommended to have
body radii and linear coordinates filled in.
Information about invalid values (NoData values) and minimum and maximum values are useful
to correctly display the image dynamics.

>>>>

\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay,
Orsay, France}
\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center,
Flagstaff, AZ, USA}
\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
CNRS, Sorbonne Universit\'es, UPMC Univ. Paris 06, Univ. Paris Diderot,
Sorbonne Paris Cit\'e, Meudon, France}
\affiliation{4}{Jacobs University, Bremen, Germany}
\affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
\affiliation{6}{Space Science Data Center - ASI, Via del Politecnico snc, 00133, Rome, Italy}
\affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.).
Planetary missions produce an impressively growing amount of heterogeneous
data which mining evolves those days from mostly visual assessment to more
automated quantitative analysis.
Interoperability and openness of data formats and processing techniques are
becoming a necessity, to avoid the risk of being unable to efficiently extract
scientific information from the data, and to guarantee the reproducibility of
the scientific results.
This paper will describe how Flexible Image Transport System (FITS) can be
used in planetary surface investigations, and how its metadata can easily be
inserted in the PDS4 metadata distribution model.
An evolution to the actual standard is proposed to adapt FITS to
planetary body description.
Some examples and implementations are then presented.
More than imposing a formal data model, FITS description for those data aims
to simplify the processing chain in planetary science, promoting
interoperability from data acquisition to final visualization.
\end{abstract}

@@

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open and flexible, largely implemented in open and efficient
processing tools.
Those reasons make FITS format appropriate for large amounts of raster data
automatic processing.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary domain could potentially
homogenize methods from acquisition, to visualization, while
giving chances to optimize data processing.

This paper is organized as follows:
\begin{enumerate}
\item{a brief description of FITS format;}
\item{a proposition for planetary surface metadata description in FITS;}
\item{dictionaries between FITS, PDS metadata description and EPN-TAP2.0
parameters;}
\item{some examples of planetary dataset in FITS;}
\item{current developments for making FITS usable in planetary sciences.}
\end{enumerate}

@@

\subsection{FITS Format}
FITS is an open digital standard, defined for data acquisition and archiving
in astronomical observatories in the late 70's, and it is used for
spatial telescope data.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

FITS is already able to propose standard formatting for some data products
quite common by now in planetary surface investigations.
In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data,
vector, table data, etc.) in the same file, each with correspondent metadata,
as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
or hyperspectral cubes with geometry information (e.g. from CRISM
\citep{crism} or OMEGA \citep{omega} instruments).
FITS has been already chosen to distribute, e.g.,
Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by
the IAU Working Group on Cartographic Coordinates and Rotational Elements
(WGCCRE).
This group reports triennially \citep{wgccre} on the preferred rotation rate,
spin axis, prime meridian, and reference surface for planets and satellites
ensuring that cartographic endeavors are effectively comparable.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
project an extension to FITS metadata (GeoFITS) has been
proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
taking into account the WGCCRE recommendations \citep{MarmoLPSC}.

@@

FITS World Coordinate System
representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html}
simplifies the spatial coordinate description with respect to historical
terrestrial references.
There is no need for oblique projection definitions
\citep{calagreisenI,Snyder1987} as projection parameters
(projection center or reference point)
describe the difference between, e.g., the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
This simplification is effective when a spherical local approximation
is applied on planetary imagery.
This is particularly the case with growing image resolutions and
increasing computing capabilities: spherical projections with a local
radius definition are easier to reproject from one system to another, than
from or to a global spheroidal datum (see for example how the HiRISE team
has chosen to distribute HiRISE images).
In addition, there is no need for East or West positive longitude
definitions as the pixel conversion in world coordinates is set by an
oriented matrix.
If needed, FITS does offer the possibility to fully describe the projected
coordinate system using several alternative WCS definitions (meters, degrees,
East and West longitude, etc.)\citep{greisencalaII}.

Table \ref{table:proj} summarizes the FITS projection codes and their
correspondent PDS4 projection names (see table A.1. in \citet{calagreisenI}
for details).

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: their correspondent in PDS4 vocabulary are
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
\end{tabular}
\end{table}

@@

Table \ref{table:wcskeys} describes FITS WCS keywords and their recommended use in
planetary surface context.

\begin{sidewaystable}
\label{table:wcskeys}
\caption{WCS FITS keywords in the Planetary Surface context}
\centering
\begin{tabular}{l l l l l}
\hline
Keyword & Type & Definition & PDS4 & EPN-TAP \\
\hline
RADESYS	& string & Name of the inertial reference frame with respect & & \\
& & to which the coordinate system is defined & & \\
& & (defaulted to \texttt{'ICRS'}) & & \\
EQUINOX & real & The value field contains the equinox in years for the celestial & & \\
& & coordinate system in which positions are expressed & & \\
& & (defaulted to \texttt{'J2000'}) & & \\
WCSAXES & integer & Number of axes in the WCS description & & \\
& & The value of WCSAXES may exceed the number of raster pixel axes & & \\
WCSNAME & & & & \\
CRPIXn & real & & \\
CTYPEn & string & & \\
CRVALn & real & & \\
CDi\_j & real matrix & & \\
PCi\_j & real matrix & & \\
CDELTn & real & & \\
CUNITn & string & & \\
CNAMEn & string (up to 68 characters) & & \\
\hline
\end{tabular}
\end{sidewaystable}

\begin{table}
\label{table:ctype}
\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.}
\centering
\begin{tabular}{l l}
\hline
Body & CTYPE$^{a}$ \\ 
\hline
Moon & SE \\
Venus & VE \\
Mars & MA \\
Jupiter & JU \\
Saturn & SA \\
Uranus & UR \\
Neptune & NE \\
Planetary bodies & B \\
\hline
\multicolumn{2}{l}{$^{a}$From \citet{calagreisenI}}
\end{tabular}
\end{table}

@@

\subsubsection{Correspondence Between Reserved FITS Keywords and PDS4 Standard Reference}
\label{sssec:stdfitspds4}
The correspondence between mandatory and reserved FITS keywords (as defined in
the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
and PDS4 attributes, is described here.
\begin{enumerate}
\item{\texttt{BITPIX}.
It specifies the number of bits that represent a data value in
the associated data array.
PDS4 uses the attribute \texttt{data\_type}: correspondent 
values depend on the associated object class.}

@@

\item{\texttt{NAXISn}.
It contains a non-negative integer representing the number of
elements along axis \texttt{n} of a data array.
The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
the count of the number of elements along an array.}

\item{\texttt{BUNIT}.
It contains a string, describing the physical units in which the quantities
in the array are expressed.
The units of all FITS header keyword values should conform with the recommendations
in the IAU Style Manual.
For angular measurements given as floating point values, degrees are the recommended
units (with the units, if specified, given as 'deg').
PDS4 specifies in the class Element (of a vector, array, or any other class
containing Elements), a \texttt{unit\_id}} (unit abbreviation) and a \texttt{unit}
(unit long name).

\item{\texttt{BSCALE}.
It contains a floating point number representing the coefficient of the linear term
in the scaling equation $physical\_value = BZERO + BSCALE \times array\_value$,
the ratio of physical value to array value at zero offset.
For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
attribute. Both FITS and PDS4 defaulted this value to $1$.}

\item{\texttt{BZERO}.
It contains a floating point number representing the constant term in the
scaling equation $physical\_value = BZERO + BSCALE \times array\_value$
For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
attribute. Both FITS and PDS4 defaulted this value to $0$.}

\item{\texttt{BLANK}.
This keyword must be used only in integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.
On floating point data the IEEE NaN must be used to represent undefined values.  
\texttt{PDS4 Special\_Constants?}}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UT
coordinates.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
they must always be formatted in UTC and Julian Days.}
% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
% 
% > Julian Days
% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
% 
% ^.

\item{\texttt{INSTRUME}.
It contains a character string identifying the instrument used to acquire the data.
PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally corresponds
to this FITS keywords.
EPN-TAP provides with an \texttt{instrument\_name} parameter.}

\item{\texttt{TELESCOP}.
It contains a character string identifying the telescope used to acquire the data.
The PDS4 \texttt{Instrument\_Host} class provides the description of the physical object
mounting the instrument. Its \texttt{Name} attribute should correspond to the FITS keyword.
PDS4 also provides for the definition of a \texttt{Telescope} class, in case of
ground-based observations.
In the EPN-TAP scheme this will defined the \texttt{instrument\_host\_name} parameter.
\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}

@@

%
%\item{\texttt{ORIGIN}.
%It contains string identifying the organization or institution responsible for
%creating the FITS file.
%}
%
%\item{\texttt{AUTHOR}}

\item{\texttt{REFERENC}.
It contains a character string citing a reference where the data are published.
It is recommended that either the Astrophysics Data System bibliographic databases\footnote{http://adswww.harvard.edu/}
bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
be used.
PDS4 \texttt{External\_Reference} class 
EPN-TAP \texttt{bib\_reference} has exactly the same meaning and use than the FITS keyword.}

@@

\subsubsection{Visualization of Global and Projected Maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
and GeoTIFF are completely interchangeable.
As an example we have downloaded some global maps from USGS
Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
We used ds9 to display the FITS image and were able to visualize coordinates in meters
as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
description (see figure \ref{fig:ds9qgis}, left side).
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9;
right: GeoTIFF raster superposed to the virtual
description of the FITS raster.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

Assuming projected products distributed in FITS format, a quick method to add GIS support for
FITS images is to use a GDAL-supported detached header.
GDAL Virtual Header files simply describe the internal structure of a FITS image, but
also the map projection and body size in a standardized "well known text"
(WKT) projection string.
As a result, the FITS raster loaded into QGIS thanks to its correspondent GDAL Virtual Header
is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).

In order to straightforward load FITS files in GIS software, it is recommended to have
body radii and linear coordinates filled in.
Information about invalid values (NoData values) and minimum and maximum values are useful
for correct display.
